---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# TimeTeller

<!-- badges: start -->
[![Version](https://img.shields.io/badge/version-2.0.0-blue)]()
[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
<!-- badges: end -->

TimeTeller is a machine learning tool for estimating **circadian clock phase** and **clock dysfunction (Θ)** from a single transcriptomic sample. It models the multi-dimensional state of the circadian clock by projecting gene expression onto local principal component spaces fitted to known circadian time series, then evaluates multivariate normal likelihoods across the 24-hour cycle to infer the most likely phase and quantify how well-ordered the clock signature is.

TimeTeller works with both **microarray** and **RNA-seq** data and supports any organism for which circadian time-course training data is available.

## Key features

- **Single-sample phase estimation** — no time-course design required for test samples
- **Clock dysfunction score (Θ)** — quantifies deviation from a healthy circadian programme
- **Automated gene selection** — population cosinor-based ranking with MRL, amplitude CV, and composite weighting via `choose_genes_tt()`
- **Wasserstein geodesic interpolation** — optional geometry-aware covariance interpolation that guarantees positive-definiteness (Mallasto et al. 2020)
- **Cross-validation framework** — leave-one-out CV with group-level prediction error and Θ diagnostics
- **Publication-ready visualisation** — consistent ggplot2 theme with colourblind-safe palettes
- **Parallel computation** — optional multi-core support for likelihood and theta calculations

## Installation

Install the development version from GitHub:

```r
# install.packages("devtools")
devtools::install_github("VadimVasilyev1994/TimeTeller-v2")
```

## Quick start

```{r example, results='hide', message=FALSE, warning=FALSE}
library(TimeTeller)

# Train on oral mucosa microarray data (Bjarnason et al.)
# 10 individuals, 6 time points each, 4h apart
tt <- train_model(
  exp_matrix = bjarn_data$expr_mat,
  genes      = bjarn_data$probes_used,
  group_1    = bjarn_data$group,
  time       = bjarn_data$time,
  log_thresh = -5
)
```

```{r train-results}
# Training results
cat("Samples:", nrow(tt$Train_Data$Results_df), "\n")
cat("Theta range:", round(range(tt$Train_Data$Results_df$Theta), 3), "\n")
cat("MAE (hours):", round(mean(abs(tt$Train_Data$Results_df$Pred_Error)), 2), "\n")
```

## Standard workflow

A typical TimeTeller analysis follows these steps:

### 1. Train the model

```r
tt <- train_model(
  exp_matrix = training_expression,   # genes x samples matrix
  genes      = gene_names,            # character vector of gene/probe IDs
  group_1    = individual_ids,        # grouping variable (e.g. individual)
  time       = sample_times,          # known zeitgeber/clock times
  log_thresh = -5                     # log-likelihood threshold
)
```

### 2. Select the log threshold

```r
lt_df <- choose_logthresh(tt, max_log = 0, min_log = -10, by_step = -0.5)
choose_logthresh_plot(lt_df)
```

### 3. Test on new samples

```r
tt <- test_model(
  tt,
  exp_matrix   = test_expression,
  test_group_1 = test_ids,
  test_time    = test_times,       # known times (for validation) or NA
  log_thresh   = -5
)
```

### 4. Assess results

```r
# Phase estimates and dysfunction scores
View(tt$Test_Data$Results_df)

# Visualise
plotPCs_test(tt)
exprs_vs_PredTime_plot(tt, genes = gene_names[1:4], theta_thresh = 0.3)
```

### 5. Cross-validation

```r
cv_res <- train_cv(
  group_to_leave_out = 'group_1',
  exp_matrix = training_expression,
  genes = gene_names,
  group_1 = individual_ids,
  time = sample_times,
  log_thresh = -5,
  test_grouping_vars = 'group_1'
)

plot_cv_res(cv_res)
plot_deviation_cv_original(cv_res)
```

## Gene selection

TimeTeller includes an automated gene selection procedure based on population cosinor analysis:

```r
tt <- choose_genes_tt(tt, pval_cutoff = 0.05)
View(tt$Rhythmicity_Results)
```

Genes are ranked by a weighted composite of p-value, R², phase consistency (mean resultant length), and amplitude reliability (CV of relative amplitude). The weights are customisable via the `rank_weights` argument.

## Covariance interpolation

By default, multivariate normal parameters are interpolated between training time points using periodic splines (`cov_path = 'spline'`). TimeTeller v2 also supports **Wasserstein (Bures) geodesic interpolation**, which preserves the positive-definite geometry of covariance matrices:

```r
tt <- train_model(
  ...,
  cov_path = 'wasserstein'
)
```

To assess the proportion of interpolated covariances requiring positive-definiteness correction under spline interpolation, use the `diagnose_pd` argument:

```r
tt <- train_model(..., diagnose_pd = TRUE)
# nearPD diagnostic: X / Y covariance matrices required correction (Z%)
```

## Bundled datasets

| Dataset | Description | Organism | Platform |
|---------|-------------|----------|----------|
| `bjarn_data` | Oral mucosa, 10 individuals × 6 time points | Human | Microarray |
| `panda_data` | Baboon multi-tissue circadian atlas | Baboon | RNA-seq |

## Citation

If you use TimeTeller, please cite:

> Vlachou D, Veretennikova MA, Stringari C, Vasilyev V, et al. TimeTeller: a tool to probe the circadian clock as a multigene dynamical system. *PLOS Computational Biology*. 2024;20(4):e1011779.

## Related publications

- Vlachou et al. (2024) TimeTeller: a tool to probe the circadian clock as a multigene dynamical system. *PLOS Computational Biology*, 20(4):e1011779.
- Bjarnason et al. (2001) Circadian expression of clock genes in human oral mucosa and skin. *The American Journal of Pathology*, 158(5):1793-1801.

## License

GPL (>= 3)
